---
title: "analyses_rdms"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Plotting RDMs as heat maps**
Visual inspection is very important to understand and interpret similarity data. R offers many packages and options to visualize heat maps. Here are just a few examples. HINT: To properly read the plot, the reader must know what the colour scale represents (legend).

## Load data
```{r}
# Load data (RDM)
# Header as F because you don't want first row of your data to become colnames
df1<-read.csv("rdm_exp1_freesorting.csv",header=F)
df2<-read.csv("rdm_exp2_acoustics.csv",header=F)
df3<-read.csv("rdm_exp3_semantics.csv",header=F)
```

```{r}
# Colnames and rownames for your matrix
animacy_order <- c(
    'LaughingAdultFemale',
    'LaughingAdultMale',
    'LaughingChild',
    'LaughingGroup',
    'CryingAdultFemale',
    'CryingAdultMale',
    'CryingChild',
    'ChildrenPlaying',
    'CocktailParty',
    'CrowdCheering',
    'CrowdApplause',
    'CoughingAdultFemale',
    'CoughingAdultMale',
    'Sneezing',
    'Snoring',
    'ThroatClearing',
    'Footsteps',
    'CatHissing',
    'CatMeowing',
    'CatPurring',
    'BirdSong',
    'BirdSquawk',
    'BirdsFlyingOff',
    'DogBarking',
    'DogGrowling',
    'DogHowling',
    'HorseGalloping',
    'HorseNeighing',
    'HorseSnorting',
    'CowMoo',
    'InsectBuzzing',
    'InsectsStridulation',
    'CarHorn',
    'CarSkidding',
    'CarStarting',
    'TrainBreaks',
    'TrainSignal',
    'TrainWagons',
    'Helicopter',
    'Propeller',
    'JetPassing',
    'DrillingPneumatic',
    'Drilling',
    'SawingManual',
    'Sawing',
    'Hammer',
    'Fire',
    'Rain',
    'River',
    'SeaWaves',
    'Thunder',
    'Waterfall',
    'Wind'
)
```

```{r}
# Rename colnames and rownames in your matrix with the proper sound names
colnames(df1) <- animacy_order
rownames(df1) <- animacy_order
colnames(df2) <- animacy_order
rownames(df2) <- animacy_order
colnames(df3) <- animacy_order
rownames(df3) <- animacy_order
```


## Heatmap with stats package
R documentation: https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/heatmap
```{r}
library(stats)
```

```{r}
# Represent your data as numeric matrix
df1_mat <- data.matrix(df1, rownames.force = NA)
df2_mat <- data.matrix(df2, rownames.force = NA)
df3_mat <- data.matrix(df3, rownames.force = NA)
```

```{r}
# Plot with clustering dendrigrams
stats::heatmap(df1_mat, keep.dendro = T, scale='none')
stats::heatmap(df2_mat, keep.dendro = T, scale='none')
stats::heatmap(df3_mat, keep.dendro = T, scale='none')
```

```{r}
# Plot without clustering dendrograms
stats::heatmap(dfmat_mat, keep.dendro = F, Colv=NA, Rowv=NA, scale='none')
```


## Save your plot
(1) This will save a jpeg file in your current directory. The resolution of the image by default will be 480x480 pixels.
```{r}
jpeg(file="name_your_plot_1.jpg")
stats::heatmap(dfmat_mat)
dev.off()
```

(2) This will save a png file in your current directory. You can adjust the size of your image.
```{r}
png(file="name_your_plot_2.png", width=600, height=350)
stats::heatmap(dfmat_mat)
dev.off()
```

(3) This will save a png file in your current directory. You can adjust the resolution of your image.
```{r}
png(file="name_your_plot_3.png", width=3.25, height=3.25, units = "in", res = 1200, pointsize = 4)
stats::heatmap(dfmat_mat)
dev.off()
```


## Applying fancy colors to your heat map
Palettes: http://applied-r.com/rcolorbrewer-palettes/
```{r}
library(RColorBrewer)
mycolors <- colorRampPalette(brewer.pal(8, "PiYG"))(25)
stats::heatmap(dfmat_mat, col = mycolors)
```


## Heat map with gplots package
```{r}
library("gplots")
heatmap.2(dfmat_mat, scale = "none", col = bluered(100), 
          trace = "none", density.info = "none")
```


## Heat map with pheatmap package
```{r}
library("pheatmap")
pheatmap(dfmat, cutree_rows = 4)
```
